# Factors affecting correlation coefficients {#factors}

```{r, warning = FALSE, echo = FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(patchwork))
```

## Overview
Observed values of correlation coefficients are affected by a number of factors to be examined here: 

| Factor               | Consequences               | Fix?                                                            |
|----------------------|----------------------------|-----------------------------------------------------------------|
| Restriction of range | Usually reduce *r<br>*     | “Correction for restriction of range"                           |
| Outliers             | Reduce or increase <br>*r* | Ideally, detect a priori                                        |
| Nonlinearity         | Misleading *r*             | Ideally detect a priori (later lecture)                         |
| Dichotomization      | Usually reduce *r*         | Don't dichotomize (corrections avalible but not discussed here) |

## Restriction of range
Within a given law school, there is a near-zero correlation between undergraduate GPA and LSAT scores. This is an instance f restriction of range affecting the correlation. 

Suppose we are interested in the realtionship between $X_1$ (UGPA) and $Y$ (LSAT). However, individuals measured on $X_1$ are **selected** to be in the sample (admitted to law school) only if their score on $X_1$ is sufficiently high, exceeding some threshold. 

In such situation, the range of $X_1$ is said to be *restricted*

It is useful ti distinguish between the correlation when selection had occured vs. the correlation if no selection had occurred. The difference can be seen in the following representation of a scatterplot: 

```{r, echo=FALSE, fig.width = 6, fig.height = 3}
ggplot(faithful, aes(waiting, eruptions)) +
  geom_point() +
  stat_ellipse() +
  labs(x = "X_1 (UGPA Score)", y = "Yy (LSAT") +
  geom_vline(xintercept = 80, color = "red") +
  theme_classic()
```

If no selection occurs, the relationship between $X_1$ and $Y$ is represented by the full ellipse. If selection occursm the relationship is represented by only the section of the ellipse beyoond the selection threshold on $X_1$. It is evident that the correlation would be greater without selection. 

This phenomenon impacts interpretation of correlations in s selection situation. After selecting for $X_1$, the observed value of $r_{Y1}$ may suggest a weaaker relationship then is actually present. 

### Correction for restriction of range

## Outliers

## Nonlinearity

## Dichotomization
